% advent25.tex
% Hero day: December 25, 2025 – Vacuum: the invisible medium of all fields

\documentclass[a4paper,10pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{titlesec}
\usepackage{lettrine}
\usepackage{eso-pic}
\usepackage{geometry}
\usepackage{multicol}
\usepackage{lmodern}

\geometry{margin=2.0cm}

\input{advent-layout}

\begin{document}

\AdventPageBackground
\AdventAuthor

\AdventSheetTwoCol
  {December 25, 2025} % #1 Date
  {}                 % #2 unused
  {Vacuum: the invisible medium of all fields} % #3 Main title
  {How $\langle H\rangle$, radii and $Y_S$ structure the ``empty'' state} % #4 Subtitle
  {Vacuum is not empty. In the octonionic model, it is a specific
   configuration of the internal structure: a vacuum expectation value
   $\langle H\rangle$ in $H_3(\mathbb{O})$, a choice of radius spectrum
   $(a_0,b_0,c_0)$ and an electroweak scale $Y_S$. Masses, couplings and
   even vacuum energy arise as properties of this structured medium. The
   vacuum is the invisible actor that sets the stage for all visible
   physics.} % #5 Key Insight
  { % #6 body

\section*{Vacuum as a configuration, not as nothing}

\AdventInitial{I}{n} quantum field theory, the vacuum is the lowest-energy
state of the system. In the octonionic setting, and within the unified master action
$S[D,\Psi]$ introduced on December 24, this state is described
by:

\begin{itemize}
  \item a vacuum configuration $\langle H\rangle \in H_3(\mathbb{O})$,
  \item a radius spectrum $(a_0,b_0,c_0)$ of the operator $R$,
  \item an electroweak scale $Y_S$ defined by the minimum of a potential.
\end{itemize}

Symbolically, one may write the internal vacuum data as

\[
  \mathcal{V}_{\text{int}} = \big(\langle H\rangle; a_0,b_0,c_0; Y_S\big).
\]

Fields are then fluctuations around this background; their masses and
interactions are determined by how they probe this structured medium.

\section*{Masses and couplings from the vacuum}

Several quantities introduced earlier in the calendar are now recognised
as vacuum properties:

\begin{itemize}
  \item Fermion masses: eigenvalues of
        $M = y\,\Pi(\langle H\rangle)$.
  \item Gauge couplings: norms of rotor commutators evaluated in the
        vacuum geometry, e.g.\ $\alpha$, $\alpha_s$, $\sin^2\theta_W$.
  \item Scales: exponentials of $(a_0,b_0,c_0)$ define preferred energies.
\end{itemize}

What looks like a zoo of independent parameters from a naive perspective
turns out to be a small set of derived quantities from
$\mathcal{V}_{\text{int}}$.

\section*{Vacuum energy and dark energy}

The cosmological constant $\Lambda$ can be understood as a contribution
to the effective action coming from the spectrum of an appropriate Dirac
operator $D$ and the vacuum configuration. In a spectral-action viewpoint,

\[
  S_{\text{spec}} = \mathrm{Tr}\, f(D^2/\Lambda^2)
\]

contains a leading term that behaves like a cosmological constant (consistent with the gravitational and defect terms in the master
action $S[D,\Psi]$ discussed on the previous day). In the
octonionic setting:

\begin{itemize}
  \item The internal structure encoded in $\langle H\rangle$ and $R$
        influences the spectrum of $D$.
  \item The corresponding vacuum energy becomes a property of the same
        medium that gives rise to masses and couplings.
\end{itemize}

This suggests a picture in which dark energy is not an exotic fluid but
an aspect of the spectral vacuum defined by the octonionic geometry.

\section*{Vacuum as an invisible actor}

It is tempting to think of particles and forces as the main characters of
physics. In this model, the vacuum plays a more central role:

\begin{itemize}
  \item It determines the scales at which different phenomena occur.
  \item It shapes the spectrum of excitations (masses).
  \item It contributes to the energy content of the universe.
\end{itemize}

Most of the ``mystery parameters'' of particle physics are reinterpreted
as properties of a single invisible actor: the structured vacuum state.

\section*{Conceptual shift}

The conceptual shift can be summarised as follows:

\begin{enumerate}
  \item \textbf{From emptiness to medium:}
    vacuum is not the absence of structure but the most structured part
    of the theory.
  \item \textbf{From input to output:}
    constants, scales and masses are not arbitrary inputs but derived
    characteristics of $\mathcal{V}_{\text{int}}$.
  \item \textbf{From fragmentation to coherence:}
    one unified internal configuration explains many seemingly unrelated
    numbers.
\end{enumerate}

This completes the Advent narrative: from abstract octonionic algebra to
a concrete picture in which the vacuum of the universe is a highly
organised medium living in that algebra.

\small
\begin{thebibliography}{9}

\bibitem{Casimir1948}
H.~B.~G.~Casimir,
\newblock ``On the attraction between two perfectly conducting plates,''
\newblock {\em Proc.\ Kon.\ Ned.\ Akad.\ Wet.} \textbf{51}, 793--795 (1948).

\bibitem{Perlmutter1999}
S.~Perlmutter \emph{et al.},
\newblock ``Measurements of $\Omega$ and $\Lambda$ from 42 high-redshift
supernovae,''
\newblock {\em Astrophys.\ J.} \textbf{517}, 565--586 (1999).

\bibitem{Internal}
[Internal notes on vacuum structure and spectral contributions:
{\tt unified-agebra.tex; chap16\_neu.tex; appM\_neu.tex; appE\_neu.tex}.]

\end{thebibliography}
\normalsize

  } % end #6 body
  {In the octonionic model, the vacuum is a structured medium specified by
   $\langle H\rangle$, $(a_0,b_0,c_0)$ and $Y_S$. Masses, couplings and
   even dark-energy-like terms become properties of this medium rather
   than independent inputs, turning the ``empty'' state into the central
   player of the theory.} % #7 Closing

\end{document}