% advent04.tex
% December 4, 2025 – Heptagon operator: seven directions, one spectrum

\documentclass[a4paper,10pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{titlesec}
\usepackage{lettrine}
\usepackage{eso-pic}
\usepackage{geometry}
\usepackage{multicol}
\usepackage{lmodern}

\geometry{margin=2.0cm}

\input{advent-layout}

\begin{document}

\AdventPageBackground
\AdventAuthor

\AdventSheetTwoCol
  {December 4, 2025} % #1 Date
  {}                 % #2 unused
  {Heptagon operator: seven directions, one spectrum} % #3 Main title
  {From seven imaginary units to three eigenvalues $(\alpha,\beta,\gamma)$} % #4 Subtitle
  {The seven imaginary octonion units can be arranged on a heptagon that
   encodes their multiplication rules. From this data one constructs the
   \emph{heptagon operator} $H_7$, a single linear operator whose
   eigenvalues $(\alpha,\beta,\gamma)$ summarise the internal geometry.
   These three numbers are not free parameters; they are geometric
   invariants that will later feed into coupling constants and mixing
   angles. Today we introduce $H_7$ and its spectrum as a compact
   fingerprint of the sevenfold structure of $\mathbb{O}$.} % #5 Key Insight
  { % #6 body (two columns)

\section*{Seven imaginary units on a heptagon}

\AdventInitial{O}{ctonions} $\mathbb{O}$ have seven imaginary units
$e_1,\dots,e_7$. Their multiplication can be depicted on an oriented
heptagon: each directed edge (and certain chords) carries a triple
$(e_i,e_j,e_k)$ with

\[
  e_i e_j = e_k,\qquad
  e_j e_k = e_i,\qquad
  e_k e_i = e_j,
\]

and reversed order introduces a minus sign. This Fano-type diagram is more
than a mnemonic; it packages the nonassociative multiplication table into
a single geometric picture.

\section*{Constructing the heptagon operator}

From the seven units $e_i$ and the heptagon geometry, one defines a linear
operator $H_7$ on the underlying real 8-dimensional space. Schematic
example:

\[
  H_7 = \sum_{i=1}^7 c_i\,L_{e_i},
\]

where $L_{e_i}:x\mapsto e_i x$ denotes left multiplication by $e_i$, and
the coefficients $c_i$ are chosen to respect the heptagon symmetry (for
instance, equal on edges belonging to the same orbit under $G_2$).

The important facts are:

\begin{itemize}
  \item $H_7$ encodes all seven directions in a single operator.
  \item $H_7$ is constrained by $G_2$ symmetry, so its spectrum is highly
        structured.
\end{itemize}

Details of the precise coefficients are not needed on this page; they are
spelled out in the technical notes.

\section*{Spectrum: three eigenvalues $(\alpha,\beta,\gamma)$}

A key feature of $H_7$ is that its eigenspectrum is much simpler than an
arbitrary $8\times8$ matrix would allow. In favourable constructions,

\[
  \mathrm{spec}(H_7) = \{\alpha,\beta,\gamma\},
\]

with appropriate multiplicities that add up to 8. The numbers

\[
  (\alpha,\beta,\gamma)
\]

are then \emph{heptagon eigenvalues}: geometric invariants that capture
how the seven imaginary directions are arranged relative to each other.

Qualitatively:

\begin{itemize}
  \item $\alpha$ may correspond to an eigen-subspace aligned with a
        particular family of edges.
  \item $\beta$ and $\gamma$ correspond to other symmetry-related
        subspaces.
\end{itemize}

The crucial point: instead of seven arbitrary numbers, we get only three
characteristic eigenvalues.

\section*{Why $(\alpha,\beta,\gamma)$ matter for physics}

Later in the calendar, $(\alpha,\beta,\gamma)$ will reappear in several
contexts:

\begin{itemize}
  \item In the construction of the \emph{radius operator} $R$ and its
        spectrum $(a_0,b_0,c_0)$ (5 December).
  \item In geometric expressions for coupling constants, such as the
        fine-structure constant $\alpha$ and the strong coupling
        $\alpha_s$.
  \item In defining angles between rotor directions that enter the
        Weinberg angle $\theta_W$.
\end{itemize}

In other words, the heptagon eigenvalues are a compact \emph{seed} from
which many later observables can be grown. They serve as a small set of
internal numbers that eventually manifest as physical constants.

\section*{Heptagon operator within the operator toolbox}

On the second Advent Sunday (7 December), the calendar will present a full
\emph{operator toolbox}:

\begin{itemize}
  \item heptagon operator $H_7$ with eigenvalues $(\alpha,\beta,\gamma)$,
  \item radius operator $R$ with radii $(a_0,b_0,c_0)$,
  \item sign/signature operators,
  \item rotors (antisymmetric generators of forces),
  \item compressors (symmetric mass and mixing operators).
\end{itemize}

The heptagon operator is the first piece of this toolbox to appear
explicitly. Its role is to turn the combinatorial data ``seven imaginary
units on a heptagon'' into spectral data that can be plugged into
operators, potentials and eventually quantitative formulas.

\section*{Conceptual gain from $H_7$}

Compared to working directly with seven basis elements $e_i$, the
heptagon-operator viewpoint offers:

\begin{enumerate}
  \item \textbf{Compression:} seven directions are summarised by three
    eigenvalues.
  \item \textbf{Invariants:} $(\alpha,\beta,\gamma)$ are invariant under
    $G_2$-compatible re-labellings of the heptagon—they are not artefacts
    of a particular basis choice.
  \item \textbf{Spectral language:} we move from basis-dependent
    multiplication tables to basis-independent spectral data, which is
    the natural language for later spectral geometry.
\end{enumerate}

Thus, $H_7$ is the first place where the octonionic multiplication table
starts to look like something a physicist would recognise as ``spectral
parameters''.

\small
\begin{thebibliography}{9}

\bibitem{Baez2002}
J.~C.~Baez,
\newblock ``The octonions,''
\newblock {\em Bull.\ Amer.\ Math.\ Soc.} \textbf{39}, 145--205 (2002).

\bibitem{Furey2016}
C.~Furey,
\newblock ``Charge quantization from a number operator,''
\newblock {\em Phys.\ Lett.\ B} \textbf{742}, 195--199 (2015).

\bibitem{Internal}
[Internal notes on the heptagon operator and its spectrum:
{\tt chap03\_neu.tex; appK\_neu.tex; oktonionen-basen.tex}.]

\end{thebibliography}
\normalsize

  } % end #6 body
  {The heptagon operator $H_7$ compresses the seven imaginary directions
   of $\mathbb{O}$ into three eigenvalues $(\alpha,\beta,\gamma)$. These
   invariants are the first internal numbers that will later reappear in
   couplings, scales and mixing angles.} % #7 Closing

\end{document}