% advent13.tex
% December 12, 2025 – Top quark: heaviest fermion as vacuum anchor

\documentclass[a4paper,10pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{titlesec}
\usepackage{lettrine}
\usepackage{eso-pic}
\usepackage{geometry}
\usepackage{multicol}
\usepackage{lmodern}

\geometry{margin=2.0cm}

\input{advent-layout}

\begin{document}

\AdventPageBackground
\AdventAuthor

\AdventSheetTwoCol
  {December 13, 2025} % #1 Date
  {}                 % #2 unused
  {Top quark: heaviest fermion as vacuum anchor} % #3 Main title
  {Why $m_t \approx 173\,$GeV sits next to the electroweak scale} % #4 Subtitle
  {The top quark mass $m_t \approx 173\,$GeV is not an outlier but the
   largest eigenvalue of the mass map $\Pi(\langle H\rangle)$ associated
   with the vacuum configuration $\langle H\rangle$ in the Albert algebra
   $H_3(\mathbb{O})$. It acts as an \emph{anchor} for the electroweak
   scale: the same structure that fixes the vacuum also singles out the
   top as the heaviest fermion. Without such an anchor, the electroweak
   scale would be unstable.} % #5 Key Insight
  { % #6 body

\section*{Masses from the mass map $\Pi(\langle H\rangle)$}

\AdventInitial{I}{n} the octonionic model, fermion masses are not free
Yukawa coefficients but eigenvalues of a linear map

\[
  \Pi : H_3(\mathbb{O}) \longrightarrow \mathrm{End}(\mathbb{R}^8),
\]

evaluated at the vacuum configuration $\langle H\rangle \in H_3(\mathbb{O})$.
The physical mass matrix is schematically

\[
  M = y\,\Pi(\langle H\rangle),
\]

with $y$ a universal coupling. Its eigenvalues

\[
  m_i = \lambda_i\big(\Pi(\langle H\rangle)\big)
\]

are the fermion masses. Among these, the largest eigenvalue is identified
with the top quark mass $m_t$.

\section*{The top as the maximal eigenvalue}

The spectrum of $\Pi(\langle H\rangle)$ is strongly constrained:

\begin{itemize}
  \item $\langle H\rangle$ is not arbitrary; it minimises a Jordan
        potential $V_J(H)$ under symmetry constraints.
  \item $\Pi$ respects the $F_4$ symmetry of the Albert algebra.
  \item The resulting eigenvalues come in structured patterns rather than
        random numbers.
\end{itemize}

In this setting, the top quark mass appears as

\[
  m_t = \max \mathrm{spec}\big(\Pi(\langle H\rangle)\big),
\]

the largest eigenvalue in the relevant sector. This is not just a
numerical statement; it has a stability interpretation.

\section*{Vacuum anchoring at the electroweak scale}

The electroweak scale $Y_S$ is fixed by the minimum of an internal
potential. Schematically,

\[
  Y_S^2 = -\frac{\mu^2}{2(\lambda+\kappa c)},
\]

with $\mu^2,\lambda,\kappa,c$ determined by the symmetry atlas. The
vacuum configuration $\langle H\rangle$ sits at this minimum, and
$\Pi(\langle H\rangle)$ inherits that structure.

The largest eigenvalue $m_t$ then behaves as an \emph{anchor}:

\begin{itemize}
  \item If $m_t$ were much smaller, the curvature of the potential near
        the minimum would change, destabilising the electroweak scale.
  \item If $m_t$ were much larger, the same structure would shift the
        position of the minimum, again spoiling the observed scale.
  \item The observed value $m_t\approx 173\,$GeV lies naturally next to
        the electroweak scale, reflecting the shared origin in
        $\langle H\rangle$.
\end{itemize}

\section*{Top quark versus lighter fermions}

Lighter fermions (electron, muon, tau; light quarks) correspond to
smaller eigenvalues of the same map. Their smallness is explained by the
geometry of $\langle H\rangle$ in the symmetry atlas:

\begin{itemize}
  \item Some directions in $H_3(\mathbb{O})$ generate large eigenvalues,
        associated with heavy fermions.
  \item Other directions generate exponentially suppressed eigenvalues,
        associated with light fermions (as discussed for neutrinos on
        15 December).
\end{itemize}

The top quark is simply the fermion whose eigenvector aligns best with
the ``steep'' direction of the potential at the vacuum point.

\section*{Conceptual picture}

In the usual Standard Model narrative, Yukawa couplings are free
parameters, and the large top Yukawa is a brute fact. In the octonionic
picture:

\begin{enumerate}
  \item The vacuum configuration $\langle H\rangle$ is fixed by internal
    geometry and potential minimisation.
  \item The mass operator $\Pi(\langle H\rangle)$ is uniquely determined
    by this vacuum and the symmetry atlas.
  \item The top quark emerges as the strongest-coupled mode to this
    vacuum, i.e.\ the maximal eigenvalue.
\end{enumerate}

The hierarchy ``$m_t$ heavy, others light'' is no longer a collection of
independent choices but a single structural statement about one operator
evaluated at one point in the Albert algebra.

\small
\begin{thebibliography}{9}

\bibitem{CDFD02014}
CDF and D0 Collaborations,
\newblock ``Combination of CDF and D0 results on the mass of the top quark,''
\newblock {\em Phys.\ Rev.\ D} \textbf{89}, 072001 (2014).

\bibitem{JordanVNW1934}
P.~Jordan, J.~von Neumann and E.~Wigner,
\newblock ``On an algebraic generalization of the quantum mechanical
formalism,''
\newblock {\em Ann.\ Math.} \textbf{35}, 29--64 (1934).

\bibitem{Internal}
[Internal notes on mass maps and the role of the top quark:
{\tt unified-agebra.tex; chap10\_neu.tex; appM\_neu.tex}.]

\end{thebibliography}
\normalsize

  } % end #6 body
  {The top quark is not an accidental heavy outlier. It is the largest
   eigenvalue of the mass map $\Pi(\langle H\rangle)$ and thus an anchor
   of the electroweak vacuum. The same internal structure that fixes the
   electroweak scale also singles out the top as the heaviest fermion.} % #7 Closing
   
\end{document}