% advent29.tex
% December 20, 2025 – Dark matter as a shadow sector of compressors

\documentclass[a4paper,10pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{titlesec}
\usepackage{lettrine}
\usepackage{eso-pic}
\usepackage{geometry}
\usepackage{multicol}
\usepackage{lmodern}

\geometry{margin=2.0cm}

\input{advent-layout}

\begin{document}

\AdventPageBackground
\AdventAuthor

\AdventSheetTwoCol
  {December 19, 2025} % #1 Date
  {}                 % #2 unused
  {Scales as attractor fixed points} % #3 Main title
  {Why nature ``chooses'' exactly three stable scales} % #4 Subtitle
  {The three fundamental scales (Planck, electroweak, QCD) are not arbitrary
   inputs but fixed points of an attractor mechanism linked to the radius
   spectrum $(a_0,b_0,c_0)$. Renormalisation-group (RG) flow drives couplings
   and masses towards these scales, which are stable under perturbations.
   Nature ``chooses'' them because they are dynamically preferred, not because
   they were inserted by hand.} % #5 Key Insight
  { % #6 body

\section*{From radius spectrum to preferred scales}

\AdventInitial{E}{arlier} in the calendar, the radius operator $R$ was
introduced with spectrum

\[
  \mathrm{spec}(R) = (a_0,b_0,c_0).
\]

Exponentials of these radii define three candidate scales,

\[
  \Lambda_i \sim e^{a_0},\quad
  \Lambda_j \sim e^{b_0},\quad
  \Lambda_k \sim e^{c_0},
\]

which were associated with the Planck, electroweak and QCD scales. So far
this was a geometric correspondence. Today we add dynamics: these scales
are not only present in the geometry, they are \emph{dynamically
selected} by RG flow.

\section*{RG flow and attractors}

In Wilsonian renormalisation, a theory is described by a point in the
space of couplings. Changing the energy scale $\mu$ corresponds to moving
along a trajectory governed by $\beta$-functions,

\[
  \frac{d g_i}{d\ln\mu} = \beta_i(\{g\}).
\]

Fixed points are defined by $\beta_i(\{g^\ast\})=0$ and play a central role:

\begin{itemize}
  \item UV fixed points control short-distance behaviour.
  \item IR fixed points control long-distance behaviour.
  \item Some fixed points are \emph{attractors}: generic RG trajectories
        are drawn towards them.
\end{itemize}

In the octonionic model, the radius spectrum $(a_0,b_0,c_0)$ enters the
$\beta$-functions in a structured way, so that three distinguished
energies become attractor scales.

\section*{Qualitative picture of the attractor mechanism}

Heuristically, one finds:

\begin{itemize}
  \item When $\mu$ is near $e^{a_0}$, RG flow of certain couplings slows
        down and stabilises; this defines a Planck-like scale.
  \item Near $e^{b_0}$, electroweak couplings and masses experience a
        similar slowdown and stabilisation.
  \item Near $e^{c_0}$, the strong coupling crosses order one and the
        theory flows towards a confined regime.
\end{itemize}

The detailed form of $\beta$-functions is model-dependent, but the
location and number of attractor scales are dictated by $(a_0,b_0,c_0)$.
Arbitrary additional scales cannot be introduced without spoiling this
structure.

\section*{Stability instead of fine-tuning}

Traditional fine-tuning problems (``Why is the electroweak scale so small
compared to the Planck scale?'') are rephrased:

\begin{quote}
  Why are there stable RG attractors at such widely separated scales?
\end{quote}

The model answers:

\begin{itemize}
  \item Because $(a_0,b_0,c_0)$ are widely separated eigenvalues of $R$.
  \item Because RG flow is structured in such a way that these eigenvalues
        generate stable fixed points.
\end{itemize}

The hierarchy is no longer a coincidence between unrelated parameters but
a built-in feature of the internal operator spectrum.

\section*{Conceptual and phenomenological implications}

Conceptually:

\begin{enumerate}
  \item \textbf{Fewer inputs:}
    instead of three independent scale parameters, one internal operator
    with three eigenvalues suffices.
  \item \textbf{Dynamical selection:}
    scales are not only present in the geometry but are dynamically
    attractive under RG flow.
  \item \textbf{Predictive structure:}
    any attempt to introduce new fundamental scales must explain how they
    fit into, or modify, the attractor pattern.
\end{enumerate}

Phenomenologically:

\begin{itemize}
  \item Small variations in UV initial conditions do not destroy the
        large-scale structure: the flow is drawn back to the attractors.
  \item This robustness is precisely what one expects from a realistic
        description of the universe: observed scales should not be
        hypersensitive to microscopic details.
\end{itemize}

\small
\begin{thebibliography}{9}

\bibitem{Wilson1971}
K.~G.~Wilson,
\newblock ``Renormalization group and critical phenomena,''
\newblock {\em Phys.\ Rev.\ B} \textbf{4}, 3174--3183 (1971).

\bibitem{Wetterich1988}
C.~Wetterich,
\newblock ``Cosmology and the fate of dilatation symmetry,''
\newblock {\em Nucl.\ Phys.\ B} \textbf{302}, 668--696 (1988).

\bibitem{Internal}
[Internal notes on scale hierarchies and attractor mechanisms:
{\tt chap16A\_neu.tex; appK\_neu.tex; konstanten-hierarchie.tex}.]

\end{thebibliography}
\normalsize

  } % end #6 body
  {The triple $(a_0,b_0,c_0)$ of radius eigenvalues does not only suggest
   three important scales; it also defines RG attractors. Planck,
   electroweak and QCD scales are dynamically preferred fixed points,
   not arbitrary external inputs.} % #7 Closing


\end{document}